dataaimsr: An R Client for the Australian Institute of Marine Science Data Platform API which provides easy access to AIMS Data Platform

### 2021-05-24 paper/paper.Rmd
removed dplyr pipe from paper Rmd
renamed statement of need, and cut down on some parts of the paper in order to address openjournals/joss-reviews#3258

# Background and Statement of need

@@

Currently, there are two AIMS long-term monitoring datasets available to be
downloaded through `dataaimsr`:

### Northern Australia Automated Marine Weather And Oceanographic Stations

Automatic weather stations have been deployed by AIMS since 1980. Most of the
stations are along the Great Barrier Reef (GBR) including the Torres Strait in
North-Eastern Australia but there is also a station in Darwin and one at
Ningaloo Reef in Western Australia. Many of the stations are located on the
reef itself either on poles located in the reef lagoon or on tourist pontoons
or other structures. A list of the weather stations which have been deployed
by AIMS and the period of time for which data may be available can be
found on the [AIMS metadata][2] webpage. **NB:** Records may not be continuous
for the time spans given.

@@

### AIMS Sea Water Temperature Observing System (AIMS Temperature Logger Program)

The data provided here are from a number of sea water temperature monitoring
programs conducted in tropical and subtropical coral reefs environments around
Australia. Data are available from approximately 80 GBR sites, 16 Coral Sea
sites, 7 sites in North West Western Australia (WA), 8 Queensland regional
ports, 13 sites in the Solitary Islands, 4 sites in Papua New Guinea and 10
sites in the Cocos (Keeling) Islands. Data are obtained from in-situ data
loggers deployed on the reef. Temperature instruments sample water
temperatures every 5-10 minutes (typically) and are exchanged and downloaded
approximately every 12 months. Temperature loggers on the reef-flat are
generally placed just below Lowest Astronomical Tide level. Reef-slope (or
where specified as Upper reef-slope) generally refers to depths 5--9 m while
Deep reef-slope refers to depths of ~20 m. For more information on the dataset
and its usage, please visit the [AIMS metadata][3] webpage.

@@

The [AIMS Data Platform API][1] points to the full metadata of each
dataset. We are currently working on ways to facilitate the 
visualisation of both datasets and their multiple features directly
through the R package. At the moment though it is only possible
to visualise summary information for the Sea Water Temperature Loggers
dataset. A similar feature for the Weather Station dataset will be 
implemented in the near future (likely late 2021)---so for now, please
refer to the online metadata to discover from where (and when) one can 
download data.

@@

The first step would be to visualise the dataset. We do this by
mapping all available sites. For example, we download the summary information
for the Sea Water Temperature Loggers dataset using the main function called
`aims_data`:

```{r, message = FALSE, warning = FALSE}
library(dataaimsr)
sdata <- aims_data("temp_loggers", api_key = NULL,
                   summary = "summary-by-series")
head(sdata)
```

Setting the argument `api_key = NULL` means that `dataaimsr` will
automatically search for the user's API key stored in `.Renviron`.
The `summary` argument here is key. It should only be flagged when the
user wants an overview of the available data. Again, this currently
implemented for the Sea Water Temperature Loggers dataset. One can
visualise `summary-by-series` or `summary-by-deployment`. The output of
`aims_data` is a `data.frame` of class `aimsdf`.

Notice that `sdata` contains a lot of information, most of which is
related to site / series / parameter ID. Each row corresponds to a
unique series, and a certain site may contain multiple series; in such
cases, series generally differ from one another by depth. The columns 
`time_coverage_start` and `time_coverage_end` are probably one of the most
valuable pieces of information. They provide the user with the window of data
collection for a particular series, which is probably crucial to decide
whether that particular series is of relevance to the specific question in
hand.

Also note that there are three columns containing the total number of 
observations in a series: `uncal_obs`, `cal_obs` and `qc_obs`, which 
respectively stand for uncalibrated, calibrated, and quality-controlled 
observations. Calibrated and quality-controlled are generally the same.
Instruments are routinely calibrated (mostly once a year) in a temperature-controlled water bath and corrections applied to the data. After
calibration, all data records are quality controlled based on the following
tests: 1) clip to in-water only data, using deployment's metadata, 2)
impossible value check: data outside a fixed temperature range (14˚C – 40˚C)
is flagged as bad data, 3) spike test: individual extreme values are flagged
as probably bad according to the algorithm presented in @morelo2014methods and
4) Excessive gradient test: pairs of data that present a sudden change in the
slope are flagged as probably bad [@toma2016acta]. If any data record fails at
least one of the tests, a QC flag equal to 2 is returned, otherwise, the QC
flag is set to 1.

One can visualise this data, for instance, by plotting them
on a map of Australia, while colouring based on the total amount of calibrated
observations (Fig. \@ref(fig:summary)).

`aimsdf` objects can be plotted using the `plot` function. For summary data
such as `sdata`, plot will always generate a map with the points around
Australia and associated regions, coloured by the number of calibrated
observations:

```{r echo = FALSE, eval = FALSE}
out <- plot(sdata, ptype = "map")
ggsave("paper/summary.png", out, width = 7, height = 4.36, units = "in",
       device = "png", dpi = 300)
```

@@

@@

```{r, message = FALSE, warning = FALSE}
aims_filter_values("weather", filter_name = "series") %>%
  head()
```

The downside is that one cannot know what time window is available
for each one of those, nor how they are nested (i.e. series /
parameter / site). In a way though the series name generally
gives that information anyway (see code output above). If knowing the 
available observation window is absolutely crucial, then as mentioned 
above the user should refer to the [online metadata][3].

@@

We recommend slicing the datasets because AIMS monitoring datasets are of very 
high temporal resolution and if one tries to download an entire series
it might take hours if not days. To slice the datasets properly, the user
needs to apply filters to their query.

>>>>

# Statement of Need

@@

Currently, there are two AIMS long-term monitoring datasets available to be
downloaded through `dataaimsr`: 1) the Northern Australia Automated Marine
Weather And Oceanographic Stations---a list of the weather stations which have
been deployed by AIMS and the period of time for which data may be available
can be found on the [AIMS metadata][2] webpage; 2) AIMS Sea Water Temperature Observing System (AIMS Temperature Logger Program)---for more information on
the dataset and its usage, please visit the [AIMS metadata][3] webpage.

@@

@@

The [AIMS Data Platform API][1] points to the full metadata of each
dataset. We are currently working on ways to facilitate the 
visualisation of both datasets and their multiple features directly
through the R package. So please consult our [on-line vignettes][35] to obtain
the most up-to-date instructions on how to navigate the different datasets.
Future versions of this package might even provide more of AIMS monitoring
datasets.

[35]: https://docs.ropensci.org/dataaimsr/articles/

@@

The first step would be to visualise the dataset. We do this by
mapping all available sites. For example, we download the summary information
for the Sea Water Temperature Loggers dataset using the main function called
`aims_data`. Setting the argument `api_key = NULL` means that `dataaimsr` will
automatically search for the user's API key stored in `.Renviron`.
The `summary` argument should only be used when the
user wants an overview of the available data---this is currently
implemented for the Sea Water Temperature Loggers dataset only. One can
visualise `summary-by-series` or `summary-by-deployment`. The output of
`aims_data` is a `data.frame` of class `aimsdf` with its own plotting
method (Fig. \@ref(fig:summary)):

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(dataaimsr)
sdata <- aims_data("temp_loggers", api_key = NULL,
                   summary = "summary-by-series")
```

```{r echo = FALSE, eval = FALSE}
library(dataaimsr)
sdata <- aims_data("temp_loggers", api_key = NULL,
                   summary = "summary-by-series")
out <- plot(sdata, ptype = "map")
ggsave("paper/summary.png", out, width = 7, height = 4.36, units = "in",
       device = "png", dpi = 300)
```

@@

For summary data such as `sdata`, plot will always generate a map with the
points around Australia and associated regions, coloured by the number of
calibrated observations.
Observations in a series can be: `uncal_obs`, `cal_obs` and `qc_obs`, which 
respectively stand for uncalibrated, calibrated, and quality-controlled 
observations. Calibrated and quality-controlled are generally the same.
Instruments are routinely calibrated (mostly once a year) in a
temperature-controlled water bath and corrections applied to the data. After
calibration, all data records are quality controlled based on the following
tests: 1) clip to in-water only data, using deployment's metadata, 2)
impossible value check: data outside a fixed temperature range (14˚C – 40˚C)
is flagged as bad data, 3) spike test: individual extreme values are flagged
as probably bad according to the algorithm presented in @morelo2014methods and
4) Excessive gradient test: pairs of data that present a sudden change in the
slope are flagged as probably bad [@toma2016acta]. If any data record fails at
least one of the tests, a QC flag equal to 2 is returned, otherwise, the QC
flag is set to 1. Please refer to our on-line [on-line vignettes][35] to learn
details about the entire structure of an `aimsdf` object.

@@

```{r, message = FALSE, warning = FALSE}
head(aims_filter_values("weather", filter_name = "series"))
```

The downside is that one cannot know what time window is available
for each one of those, nor how they are nested (i.e. series /
parameter / site). In a way though the series name generally
gives that information anyway (see code output above). If knowing the 
available observation window is absolutely crucial, then as mentioned 
above the user should refer to the [on-line metadata][3].

@@

We recommend slicing the datasets because AIMS monitoring datasets are of very 
high temporal resolution and if one tries to download an entire series
it might take a few hours. To slice the datasets properly, the user
needs to apply filters to their query.
